Variations on a Bayesian Theme: Comparing Bayesian Models of Referential Reasoning

### 2013-10-23 01_intro.tex
Latest version

Despite the RSA model's theoretical advantage in that it provides us with quantitative predictions and its empirical success, if we analyze the design of the model, we will find several choices that are not \emph{prima facie} obvious and thus need further clarification. These choices have their alternatives in the closely related game-theoretic approaches to pragmatics (e.g. \cite{Benz2007,Jager2013}), which have similar but slightly different conceptual motivations. Hence it is important to systematically compare the original RSA model with all these alternatives and test their predictions on empirical data, to gain a better understanding of the nature of human pragmatic inference.

The outline of the paper is as follows. We first introduce the simple referential communication game investigated in \cite{Frank} and the original RSA model proposed there. Next we analyze a few design choices in this model and introduce their alternatives in game-theoretic pragmatics. Finally we report on the result of an experiment similar to the original one, and compare the predictions of different models to the new data.

>>>>

Despite the RSA model's theoretical advantage in that it provides us with quantitative predictions and its empirical success, if we analyze the design of the model, we will find several choices that are not \emph{prima facie} obvious and thus need further consideration and justification. These choices have their alternatives in the closely related game-theoretic approaches to pragmatics (e.g. \cite{Benz2007,Jager2013}), which have similar but slightly different conceptual motivations. Hence it is important to systematically compare the original RSA model with all these alternatives and test their predictions on empirical data, to gain a better understanding of the relation between these models. Doing so will help illuminate the theoretical commitment and empirical implication of each design choice, which will enhance our understanding of the nature of human pragmatic inference.

The outline of the paper is as follows. We first introduce the simple referential communication game investigated in \cite{Frank} and the original RSA model proposed there. Next we analyze a few design choices in this model and introduce their alternatives in game-theoretic pragmatics, emphasizing the distinction between \emph{belief}, \emph{goal} and \emph{action} and unifying various models as different interpretations of these three components. Then we report on the result of an experiment similar to the original one \cite{Frank}, and compare the predictions of different models to the new data. Finally, we discuss the implications of the results of the model comparison, pointing out that a more complex picture of pragmatic reasoning about referential expressions awaits further exploration.

### 2013-10-23 04_experiment.tex
Latest version

The positions and colors of the objects, as well as the orders of the candidate words in the speaker conditions, were counterbalanced.

@@

There were 432 participants in the speaker conditions, 144 in each condition. The numbers of participants choosing each word in each condition are shown in Table \ref{table:speaker}. Note that when the target was the object with the unique shape (as in the first row of the table), the feature of shape (``Square'' in the example) should be the optimal utterance because the listener could uniquely identify it. Similarly when the target was the object with the unique color (the second row), the optimal utterance would be the feature of color (``Blue''). When the target was the object with both features shared, both features should be equally ambiguous because of the context's symmetric nature.

>>>>

In order to minimize the effect of the confounding factors, we counterbalanced the positions and colors of the objects, as well as the orders of the candidate words in the speaker conditions. Thus the unique color in a context could be either blue (as in Fig. \ref{exp_context}) or green and the order of the objects could be any of the permutations\footnote{Due to the scale of our preliminary experiment we did not counterbalance the shape of the objects, i.e. a context always has one square and two circles. }.

@@

There were 432 participants in the speaker conditions, 144 in each condition. The numbers of participants choosing each word in each condition are shown in Table \ref{table:speaker}. Note that when the target was the object with the unique shape (as in the first row of the table), the feature of shape (``Square'') should be the optimal utterance because the listener could uniquely identify it. Similarly when the target was the object with the unique color (the second row), the optimal utterance would be the feature of color (``Blue'' in this case). When the target was the object with both features shared, both features should be equally ambiguous because of the context's symmetric nature.

### 2013-10-23 06_conclusion.tex
Latest version

Our experiment and model comparison suggest the following: (1) On the conceptual level, 
it is helpful to clarify the distinction between belief, goal and action, and the Bayesian framework provides us with a natural way to structure these components into a formal model. By examining each component systematically, we can explicitly spell out the underlying assumptions and formulate alternative hypotheses to be tested. (2) On the empirical level, we tested the RSA model with a family of variants motivated from the game-theoretic pragmatic perspectives on communication, and in particular, we investigated the speaker's preference as well as the listener's perceptual salience. Our data showed a more intricate picture about the roles various factors play in pragmatic reasoning about referential expressions --- what is commonly known is the Gricean pragmatic components, while either side have his own private bias unknown to the other. Also, an action-based view might better reflect the goal of communication, at least in the forced-choice experimental setting.

Our work shows that careful conceptual analysis of the design choices for quantitative models can lead to a better understanding of the phenomenon and further improvement in the formal model's predictive power. Of course, more empirical data are needed for a robust pragmatic theory of people's reasoning about referential expressions.

>>>>

Our experiment and model comparison suggest the following: (1) On the conceptual level, 
it is helpful to clarify the distinction between belief, goal and action, and the Bayesian framework provides us with a natural way to structure these components into a formal model. By examining each component systematically, we can explicitly spell out the underlying assumptions and formulate alternative hypotheses to be tested. (2) On the empirical level, we tested the RSA model with a family of variants motivated from the game-theoretic pragmatic perspectives on communication, and in particular, we investigated the speaker's preference as well as the listener's perceptual salience. Our data showed a more intricate picture about the roles various factors play in pragmatic reasoning about referential expressions. There is some correlation between the speaker's and the listener's rationality,  while either side has his own bias and is unaware of the other's. Also, an action-based view might better reflect the goal of communication, at least in the forced-choice experimental setting. Understanding the relation between the forced-choice and the betting paradigms would be an important next step.

Our work shows that careful conceptual analysis of the design choices for quantitative models can lead to a better understanding of the phenomenon and further improvement in the formal model's predictive power. Of course, our preliminary experiment should only serve as a start and more empirical data are needed for a robust pragmatic theory of people's reasoning about referential expressions.

### 2013-10-23 main.tex
micha push test
model comparison speaker
ModComp nicer tables and added references
plots updated, packages added to main.tex
Latest version

@@

%\usepackage{gb4e}         % NOTE!!!   This should be the LAST  usepackage   call!!!!!!!
%\usepackage{lingmacros}
% ======== Definitions ========

@@

\begin{abstract}
Recent developments in Bayesian experimental pragmatics have received much attention, as it provides us with a quantitative model formalizing core concepts in traditional pragmatic theories and the predictions it makes fit empirical data nicely. Despite its theoretical advantage and empirical success, there are still several conceptual issues in the Bayesian Rational Speech Act (RSA) model that need further clarification. In this paper, we analyze a few design choices made in RSA that have intuitively equally plausible alternatives motivated from related game-theoretic approaches to pragmatics, and compare the original RSA model with these alternatives to empirical data. The result suggests that while RSA is better than most other variations, there is still room for improvement.
\end{abstract}

>>>>

\usepackage[table]{xcolor}

@@

%\usepackage{gb4e}         % NOTE!!!   This should be the LAST  usepackage   call!!!!!!!
%\usepackage{lingmacros}

%%% Micha's stuff
\usepackage{booktabs}
\usepackage{todonotes}

% ======== Definitions ========

@@

\begin{abstract}
Recent developments in Bayesian experimental pragmatics have received much attention, as it provides us with a quantitative model formalizing core concepts in traditional pragmatic theories and the predictions it makes fit empirical data nicely. Despite its theoretical advantage and empirical success, there are still several conceptual issues in the Bayesian Rational Speech Act (RSA) model that need further clarification. In this paper, we analyze a few design choices made in RSA, focusing on its \emph{belief}, \emph{goal} and \emph{action} components and introduce some intuitively equally plausible alternatives motivated from related game-theoretic approaches to pragmatics as different interpretations of these three components. Then we compare the original RSA model with these alternatives to new empirical data and the result suggests a more complex picture of pragmatic reasoning about referential expressions than the RSA depicts. While there is a correlation between the speaker's and the listener's pragmatic reasoning that may be seen as common rationality, either side has its own bias that the other does not take into account.
\end{abstract}
